---
title: "Module02"
author: "Caterina"
date: "2026-06-02"
output: html_document
---

#### Starting Module 2 #### 

#1.3 Memory Allocation and Environment Hygiene 
```{r}
#Housekeeping 
#Inventory every active object currently residing in sessions RAM objects 
#Purge global environment 
rm(list=ls())
#confirm that global session memory now completely vacant
objects()
```

Installing package called 'here'
```{r}
#install.packages("here")
```

#1.3 Ingesting Data into R 
Loading the primary data science framework and Excel import library 
```{r}
library(tidyverse)
library(readxl)
```
Practice Import A: Loading a standard comma-seperated plain text file:
```{r}
benthic_cover <- read.csv(here::here("data/reef_cover_log.csv"))
```
Practice Import B: Parsing a tab-seperated telemetry instrument array string 
```{r}
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))
```
Practice Import C: Targeting a specific sheet in a multi-yab Excel spreadsheet 
```{r}
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet ="Commercial_2026")
```
Try loading mangrove dataset... 
```{r}
mangrove_data <- read_csv(file=here::here("data/mangrove_survey_raw.csv"))
```
We're going to use args within read_csv to skip headers and declare missing files. 
By providing the NA chacater vector, we instruct R to search for common biological placeholder shorthands (like . or ND for "no data") and normalise them into an explicit, trackable missing indicator state - NA - right at the point of ingestion. 
```{r}
mangrove_data <- read_csv (
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5, #this skips the first 5 lines of field notes 
  na = c(".", "NA", "9999", "ND", "blank")) #convert known text alternatives to true NA 
```
# 1.5 Data frame architectures: Tibbles vs. Legacy Tables 
To see the behavioural differences directly, initialise a native tidyverse data object and enforce a comparative class transformation: 
```{r}
#Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)
```
and compare 
```{r}
#Print the old-style data frame structure to view 
print(benthic_cover_df)
#And compare with tibble alternative 
print(benthic_cover)
```
#1.6 Wrangling out ecological signals using Palmer Penguins 
To explore how data wrangling reveals information about underlying ecology, we will transition to working with a landmark built-in data set: the Palmer Penguins archive. This collection contains rigorous multi-species morphological tracking records collected at the Palmer Station long-term ecological research site in Antarctica. 

To activate the dataset and begin the exploritory audit, install the package repository and initialise the object inside your local workspace session memory: 

```{r}
#Install the data package (this will be executed in the console pane so it will not be saved) 
#install.packages("palmerpenguins")

#Load the package data into active memory 
library(palmerpenguins)
data("penguins")
```

```{r}
#Examine the structure of the dataset - always do this when loading a new dataset! 
glimpse(penguins) #tidyverse version from dplyr package 
str(penguins) #base R version 
```
Looking at the printout from the above code... the glimpse() command maps out the entire architectural anatomy of your dataset. Note the variable vector types listed within the arrow markers: 
<fct> (Factor): categorical groupings with fixed levels(e.g. species containing Adelie, Chinstrap, and Gentoo).
<dbl> (Double): continuous numeric measurements containing decimals (e.g. bill_length_mm). 
<int> (Integer): whole number variables tracking counts (e.g. body_mass_g). 
Before moving into data modification, we can also conduct a statistics overview to mao out missing observations and parameter boundaries: 
```{r}
#Generate an exploratory summary matrix 
summary(penguins)
```
The summary() printout is an immediate disgnostic tool. 







