---
title: "Module02 Data Science in R"
author: "Caterina"
date: "2026-06-11"
output: html_document
---

# Workshop 2

```{r setup, include=FALSE}
#Set options, housekeeping
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

```{r}
library(dplyr)
library(stringr)
library(lubridate)
library(tidyverse)
library(tidyr)
```

#### 2.2 Tidying Data 

Tables 

```{r}
table1
```
```{r}
table2
```
```{r}
table3
```
```{r}
#Compute rate per 10,000 
table1 %>% 
  mutate(rate = cases/population * 10000)
```
```{r}
#Compute cases per year 
table1 %>%
  count(year, wt = cases)
```
```{r}
#Visualise changes over time 
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

#### 2.3 Pivoting data to make it tidy 
Use tidyr:pivot_longer() to lengthen data and pivot_wider() to widen data. 

#### 2.4 Lengthening Datasets 
pivot_longer() makes datasets "longer" by increasing the number of rows and decreasing the number of columns, solving those common problems of data values in the variable name (eg. 2k1, wk2, wk3 etc.)
The table4 needs to be wrangled into 'long' format (represented by the table on the left) in order to satisfy the 3 rules of tidy data. 

Lets see pivot_longer() in an example using the billboard dataset records ranking songs in the year 2000. 

```{r}
billboard
```

The first step is to understand what each observation means, so we can be sure they are represented appropriatly as rows. 
In this data set, each observation is a song. The first three columns (artist, track & data.entered) are variables that describe the song. Then we have 76 columns (wk1-76) that described the rank of the song in eachweek. Here, the column names are one variable (the week) and the cell values are another (the rank). to tidy the billbaord data we will use pivot_longer(). 

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank"
  )
```

Now the data is in longform. We have the ranking of every song for every week E.g. Baby Don't Cry by 2 Pac from week 1 to week 76 with its ranking. So, we have many more pages as it is in longform. 
cols = specifies the columns you want to pivot. 
names_to = which names the variable stored in the column names, we chose to name that variable 'week'. 
values_to = which names the variable stored in the cell values that we named 'rank'. 
There are a lot of NA values throughout the weeks, so make this go away we ass values_drop_na = TRUE. 

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank", 
    values_drop_na = TRUE
  )
```

Now there are fewer rows and our data is tidy! 

#### Pivoting longer - example 
Let's see how pivoting can help reshape and see what it actually does to our data. 

```{r}
df <- tribble(
  ~id, ~bp1, ~bp2,
  "A", 100, 120,
  "B", 140, 115, 
  "C", 120, 125
)
```

Here, we have created a dataset called 'df' with 3 variables and their associated values. However, we want our new tidy dataset to have 3 variables: 
1. id (which already exists)
2. measurement (the column names) 
3. value (the cell values) 
To make this happen we need to pivot df longer... 

```{r}
df |>
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement", 
    values_to = "value"
  )
```

#### 2.5 Widening Datasets 
In less common cases, we may need to widen a dataset rather than lengthen it. Widening is the opposite of lengthening so we use the function pivot_wider(). It allows us to handle an observation if it is scattered across multiple rows. 
We'll use an example to explore pivot_wider() looking at the cms_patient_experience dataset from the Centres of Medicare and Medicaid. 

```{r}
cms_patient_experience
```

The core unit being studied is an organisation, but in this format, each organisation is spread across six rows, with one row for each measurement taken in the survey organisation. We can see the complete set of values for measure_cd and measure_title by using distinct(). 

```{r}
cms_patient_experience |>
  distinct(measure_cd, measure_title)
```

Neither of the columns name great variable names i.e. measure_cd doesn't hint at the meaning of the variable and measure_title is a long sentence containing spaces. 
pivot_wider() has the opposite interface to pivot_longer():_from) and the column name (names_from). 

```{r}
cms_patient_experience |>
  pivot_wider(
    names_from = measure_cd, 
    values_from = prf_rate
  )
```

The above output doesn't look quite right, we still seem to have multiple rows for each organisation. That's because, we also need to tell pivot_wider() which column or columns have values that uniquely identify each row, in this case those are the variables starting with "org": 

```{r}
cms_patient_experience |>
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

This output gives us what we're looking for! 

#### Pivoting Wider Example 
To understand what pivot_wider() does to our data, let's use a simple example. 
This time we have two patient id's A and B, and we have three blood pressure (bp) measurements from patient A and two from patient B. 

```{r}
df <- tribble(
  ~id, ~measurement, ~value, 
  "A", "bp1", 100, 
  "B", "bp1", 140, 
  "B", "bp2", 115, 
  "A", "bp2", 120, 
  "A", "bp3", 105
)
```

We will take the names from the measurement column using the names_from() argument and the values from the value column using the values_from() argument. 

```{r}
df |>
  pivot_wider(
    names_from = measurement, 
    values_from = value
  )
```

To start the pivoting process, pivot_wider() needs to first figure out what will go in the rows and columns. The new column names will be the unique values of measurement. 

```{r}
df |> 
  distinct(measurement) |> 
  pull()
```

This shows the variables I have, bp1 to bp3. 

```{r}
df |> 
  select(-measurement, -value) |>
  distinct()
```

This shows the ids I have, patient A and patient B. 
pivot_wider() then combines these results to generate an empty dataframe: 

```{r}
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
```

#### 2.6 Pivoting Exercises using Palmer Penguins 
Lengthening data with pivot_longer(). 
Imagine we want to create a single plot that shows the distribution of all morphometric measurements (bill length, bill depth, flipper length, and body mass) for our penguins. Currently, in tidy format, these variables are spread across four separate columns. To plot them together use facet_wrap(). 

```{r}
library(palmerpenguins)

#Create a long version of the penguins dataset. 

penguins_long <- penguins |> 
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g), 
    names_to = "measurement_type", 
    values_to = "value"
  )

# View the result 
head(penguins_long)
```

This lengthened format is exactly what ggplot2 expects for multi-panel faceting. Let's visualise these distributions simultaneously: 

```{r}
penguins_long |> 
  drop_na(value) |> 
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 50, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species", 
    x = "Measurement value", 
    y = "Frequency"
  )
```

#### Widening data with pivot_wider() 
Now, we will look at the reverse operation. Sometimes, it's much easier for a reader to look at a wide table than a long, repetitive list. 
First, lets calculate the mean body mass for each species on each island using group_by() and summarise(). 

```{r}
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)
```

This output is technically tidy and computationally useful, but it is not ideal for publication presentation. We can use pivot_wider() to pull the island names up into columns, creating a clean matrix of species by island. 

```{r}
mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )

head(mass_matrix)
```

I prefer the way this looks. 

#### 2.7 Separating and Uniting Data Tables 
In table3, we see the one column 'rate' contains two variables (cases and population). To address this, we can use separate() function, which separates one column into multiple columns wherever you designate. 

```{r}
table3
```

Now we will separate the rate column into two columns. 

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

Now, you can see there are two columns, cases and population. R knew to separate the numbers where the slash was. 
But, if you wanted to use specific characters to separate a column, you can put this into the argument: 

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/")
```

As we can see, the columns cases and population are showing as <chr> which means character, which is wrong, they are numbers, this happens because its a default of separate(). 
We can fix this by using convert = TRUE. 

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"), convert = TRUE)
```

Now the columns cases and population are <int>, as they are integer types. 

We can also seperate the last two digits of each year. This makes data less tidy, bu can be useful in some cases. 

```{r}
table3 %>%
  separate(year, into = c("century", "year"), sep =2)
```

A new column has been added, century. 

#### Using unite(). 
unite() is the inverse of separate(). We use unite() to combine multiple columns into a single column. We will use it to combine century and year columns from above. 

```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```

Nice! Now our columns are combined. We add sep="" because we dont want any separator. 

#### 2.8 Wrangling Strings & Dates 
When combining data from different sources-say, a government database, an oceanographic sensor, and handwritten field sheets-we inevitably run into formatting mismatches. A computer does not know that Nelly Bay, nelly_bay and NELLY BAY are the same location. Similarly, date formats vary wildy depending on the instrument or the researchers regional settings (e.g. DD/MM/YYYY vs. YYYY-MM-DD). 
So, we myst first standardize our text strings and date-times. The tidyverse includes two specialised packages for this... stringr for text and lubridate for dates. 

#### Standardising text with stringr. 
stringr package begins with str_. 

```{r}
library(tidyverse)

#A remarkably messy data frame of field sites 
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", "Geoffrey_Bay", "geoffrey bay")
)

#Using stringr within mutate to standardise the text: 
clean_sites <- messy_sites |> 
  mutate(
    #1. convert everything to lowercase 
    site_clean = str_to_lower(site_id),
    #2. trim any leading or trailing whitespaces (invisible spaces at the ends)
    site_clean = str_trim(site_clean),
    #3. Replace any spaces with underscores 
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
  )

print(clean_sites)
```
As we can see, all the different types of nelly bay variations and geoffery bay variations are now the same. 

#### Mastering time with lubridate 
If your dataset contains a date formatted as "25/12/2026" we would use the function dmy(). If the date is 2026-12-25, you use ymd(). 

```{r}
library(lubridate)

#Parsing different date formats. 

date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

#R now recognises these as identical Date objects. 
date_1 == date_2
```

For high-frequency marine data, such as acoustic telemetry or CTD profiles, we often need to parse exact timestamps. We do this by appending _mhs (hours, minutes, seconds) to our functions. 

```{r}
#A tibble of raw sensor data with messy character timestamp... 
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

#Converting character strings to true POSIXct datetime objects 
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)
```
#### 2.9 Relational Data: Joining Tables 
Bringing tables together based on a common identifying column, known as a key. 
We will set up two tables to demonstrate: 

```{r}
#Table 1: Biological observation data. 
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"), 
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

#Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)
```

#### The workhorse: left_ and right_join() 
The left_join() and right_join() work the same way, but in opposite directions. 
For example: left_join() keeps all the rows from your left table (observations) and matches data from your right table (site_metadata), wherever the key matches. If a match is missing, it fills in the gaps with NA. 

```{r}
#Joining metadata to our observations 
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)
```

#### The Strict Filter: inner_join()
An inner_join() only keeps rows that have complete matches in both tables. If an observation occurs at a site that isnt listed in the metadata, that observation will be dropped from the dataset. 

```{r}
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
glimpse(matched_data)
```

#### The Detective: anti_join()
The anti_join() filters your left table to show only rows that do not have a match in the right table. 
If you try a left_join() and end up with unexpected NA values, use an anti_join() to find out exactly which site codes are misspelled or missing from your metadata table. 

```{r}
#Which observations are missing from our metadata dictionary? 
missing_context <- observations |> 
  anti_join(site_metadata, by =join_by(site_code))

glimpse(missing_context)
```

#### 2.10 Handling Missing Values 
Missing data (NA) is normal in the reality of field science. Sometimes cells are deliberatly left blank to indicate zero when time is of the essence etc. 

#### Converting legacy erros with na_if() 
Some kinds of equipment have numbers that represent a missing value, e.g. like -99 or -999, as they cannot output a blank NA. If these are left in our dataset, the mean temperature calculations will be completely ruined. But, this can be fixed securley inside a mutate() using dplyr::na_if(): 

```{r}
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5) #-999 is a known sensor error code. 
)

#convert the -999 error codes to true NA values: 
fixed_logger <- logger_data |> 
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)
```

#### Replacing missing values with coalesce 
Sometimes NA can mean a known fixed number, most commonly zero, not missing data. 
E.g. a surveyor may leave a cell blank to indicate they saw zero sharks, R will import blank cell as NA. 
We can use dplyr::coalesce() inside mutate() to hunt down every NA in a specific column and replace it with a zero (or any fixed value). 

```{r}
#Simulate some count data 
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) #the NA here actually means zero sharks seen. 

#Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)
```

#### The special case of NaN (Not a Number)
Instead of, NA - we do not know what the value is, NaN means - this value is mathematicllay impossible. 
It is most common when performing a mathematical operation that has an indeterminate result, such as dividing zero by zero i.e. when calculating metrics such as Catch Per Unit Effort (CPUE). 

```{r}
cpue_data <- tibble(
  site = c("Bay_1)", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2,0))

#Calculating CPUE (catch/effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch/effort_hours)

print(cpue_calc)
```

Output in Bay_2 cpue = NaN. But can use "is.nan()" function to audit the data for mathematical errors without flagging standard missing data. 

#### Implicit missing values and the zero-data framework: complete()
In ecology, knowing where a species was not found is just as important as knowing where it was found. However, raw data almost exclusively records presences (implicit missing data). 
Example: raw catch log of coral trout species, it only lists when it was caught: 

```{r}
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"), 
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"), 
  count = c(5, 2, 8)
)
```

It's implied that Reef_2 was surveyed but no Plectopomus were found. If we calculate the mean catch across all sites, base R will tell us it is 2 (because it only knows about Reef_1), ecologically, the mean should be 1 (2 at Reef_1, and 0 at Reef_2). 
We can use tidyr::complete() to build a zero-data framework. This function finds all unique combinations of the variables you supple, explicitly creates rows for the missing combinations, and allows you to fill the resulting NAs with 0. 

```{r}
#Force inclusion of hidden zero-catch data 
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)
```

Now Reef_2 now explicitly lists Pleopardus with a count of 0. Building these frameworks is a mandatory prerequisite before reading into Generalised Linear Models (GLMs) and other analyses. 

#### The nuclear option: drop_na() 
If an NA represents a failed instrument reading, and that reading is the core response variable of your analysis, you may have no choice but to remove the entire row. We can do this cleanly using drop_na(). 

```{r}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

#Remove any row where salinity is missing 
clean_log <- sensor_log |>
  drop_na(salinity)
```

#### 2.11 Practical Exercies: Penuins, Pivots, and Relational Data 

Exercise 1: Cleaning the Messy Metadata 
Run the following code to generate the colleague's messy metadata table in your environment: 

```{r}
#Make sure packages are loaded 
#library(tidyverse)
library(palmerpenguins)

#The messy metadata provided by your colleague: 
island_metadata <- tibble( 
  island_name = c(" biscoe", "Dream", "Torgersen"), 
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"), 
  latitude = c(-64.81, -64.73, -64.76))

print(island_metadata)
```
```{r}
island_metadata_clean <- island_metadata |>
  mutate(
    station_install = dmy(station_install)
  ) |>
  mutate(
    #1. convert everything to have capital letter for island name
    island_name = str_to_title(island_name),
    #2. trim any leading or trailing whitespaces (invisible spaces at the ends)
    island_name = str_trim(island_name),
    #3. Replace any spaces with underscores 
    island_name = str_replace_all(island_name, pattern = " ", replacement = "_")
  )
```

Joining this clean metadata to the main penguins dataset and calling it penguins_spatial. 

```{r}
#Joining metadata to our observations 
penguins_spatial <- penguins |>
  left_join(island_metadata_clean, by = join_by(island == island_name))
```


