---
title: "Module02"
author: "Caterina"
date: "2026-06-02"
output: html_document
---

#### Starting Module 2 #### 

#1.3 Memory Allocation and Environment Hygiene 
```{r}
#Housekeeping 
#Inventory every active object currently residing in sessions RAM objects 
#Purge global environment 
rm(list=ls())
#confirm that global session memory now completely vacant
objects()
```

#1.3 Ingesting Data into R 
Loading the primary data science framework and Excel import library 
```{r}
library(tidyverse)
library(readxl)
library(readr)
```
Practice Import A: Loading a standard comma-seperated plain text file:
```{r}
benthic_cover <- read.csv(here::here("data/reef_cover_log.csv"))
```
Practice Import B: Parsing a tab-seperated telemetry instrument array string 
```{r}
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))
```
Practice Import C: Targeting a specific sheet in a multi-yab Excel spreadsheet 
```{r}
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet ="Commercial_2026")
```
Try loading mangrove dataset... 
```{r}
mangrove_data <- read_csv(file=here::here("data/mangrove_survey_raw.csv"))
```
We're going to use args within read_csv to skip headers and declare missing files. 
By providing the NA chacater vector, we instruct R to search for common biological placeholder shorthands (like . or ND for "no data") and normalise them into an explicit, trackable missing indicator state - NA - right at the point of ingestion. 
```{r}
mangrove_data <- read_csv (
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5, #this skips the first 5 lines of field notes 
  na = c(".", "NA", "9999", "ND", "blank")) #convert known text alternatives to true NA 
```
# 1.5 Data frame architectures: Tibbles vs. Legacy Tables 
To see the behavioural differences directly, initialise a native tidyverse data object and enforce a comparative class transformation: 
```{r}
#Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)
```
and compare 
```{r}
#Print the old-style data frame structure to view 
print(benthic_cover_df)
#And compare with tibble alternative 
print(benthic_cover)
```
#1.6 Wrangling out ecological signals using Palmer Penguins 
To explore how data wrangling reveals information about underlying ecology, we will transition to working with a landmark built-in data set: the Palmer Penguins archive. This collection contains rigorous multi-species morphological tracking records collected at the Palmer Station long-term ecological research site in Antarctica. 

To activate the dataset and begin the exploritory audit, install the package repository and initialise the object inside your local workspace session memory: 

```{r}
#Install the data package (this will be executed in the console pane so it will not be saved) 
#install.packages("palmerpenguins")

#Load the package data into active memory 
library(palmerpenguins)
data("penguins")
```

```{r}
#Examine the structure of the dataset - always do this when loading a new dataset! 
glimpse(penguins) #tidyverse version from dplyr package 
str(penguins) #base R version 
```
Looking at the printout from the above code... the glimpse() command maps out the entire architectural anatomy of your dataset. Note the variable vector types listed within the arrow markers: 
<fct> (Factor): categorical groupings with fixed levels(e.g. species containing Adelie, Chinstrap, and Gentoo).
<dbl> (Double): continuous numeric measurements containing decimals (e.g. bill_length_mm). 
<int> (Integer): whole number variables tracking counts (e.g. body_mass_g). 
Before moving into data modification, we can also conduct a statistics overview to mao out missing observations and parameter boundaries: 
```{r}
#Generate an exploratory summary matrix 
summary(penguins)
```
The summary() printout is an immediate diagnostic tool. It reports the presence of explicit missing indicators (NA) within individual biological metrics. For example, if sex or body_mass_g lists several NA counts, it flags that certain penguins escaped the researchers during that day's field sampling. 

#1.7 Foundational Grammar: slicing, filtering, sorting, and transforming 
The dplyr package provides a highly structured grammar of data manipulation built around intuitive action verbs. 

1.7.1 Isolating attributes with select() 
The select() verb allows you to slice your datasets vertically, isolating or dropping columns based on variable names: 
```{r}
#Vertically slice specific morphometric variables by explicit name. 
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics) 
```
```{r}
#Retain a continuous block of attributes using the colon operator. 
spatial_block <- select(penguins, species:island)
#Discard logistics tracking attributes while preserving everything else using the minus sign. 
clean_scientific_fields <- select(penguins, -year)
```

1.7.2 Sifting rows with filter() 
While select() targets variables vertically, filter() isolates records horizontally based on targeted conditional parameters. R evaluates every row agaidropping those that return FALSE or NA. 
```{r}
#Isolate observations belonging to a single categorical target group. 
adelie_cohort <- filter(penguins, species == "Adelie")

#Sift out individuals using continuous numerical boundary thresholds. Preserves only large penguins whose mass exceeds 4500 grams. 
heavy_penguins <- filter(penguins, body_mass_g > 4500)

#Combine multiple conditional parameters across separate attributes. Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island. 
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

#Sift records matching multiple targeting flags within an explicit set 
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```

1.7.3 Ordering sequences with arrange()
The arrange() verb alters the sorting configurations of rows within your data frame without changing individual cell values. This is essential for inspecting size hierarchies or chronologically ordering long-term environmental observation lines: 
```{r}
#Sort penguins by ascending body mass(Default setting: smallest mass first)
lighest_first <- arrange(penguins, body_mass_g)

#Sort penguins in descending sequence using the desc() layout wrapper. 
heaviest_first <- arrange(penguins, desc(body_mass_g))

#Execute nested sorting criteria: Group by species, then sort by descending bill length. 
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
```

1.7.4 Introducing the Pipe(|>)
We have been creating intermediate variables to hold the result at each stage. This is great for clarity, but it clutters our environment with a lot of objects. In R, there is a much cleaner way to chain these operations together: the pipe. 

An Example: 

```{r}
penguins_subset <- mutate(penguins, bill_ratio = bill_length_mm / bill_depth_mm)
penguins_final <- filter(penguins_subset, species == "Adelie")
```
OR we can simply write: 
```{r}
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
```
By using the pipe, the code becomes easier to read, easier to debug, and much more efficient to write. 

1.7.5 Computing new attributes with mutate() 
```{r}
#Calculate a new morphological ratio in our environment 
penguin_ratios <- penguins |> 
  mutate(body_mass_kg = body_mass_g / 1000, #convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm #bill ratio
         )

#View the newly engineered variables appended to the far-right columns. 
glimpse(penguin_ratios)
```
#1.8 Data aggregation and ecological summarisation 
Isolating and sorting individual rows or columns allows us to clean and audit a dataset, but it does not tell you the broader or underlying ecological story. To extract meaningful biological conclusions - such as determining whether morphological allometry varies by species or island group-you must compress thousands of individual observations into into explicit population summary metrics. 

This is achieved using the dual-layered combination of group_by() and summarise(). 

```{r}
#Grouping our active memory penguins by species 
grouped_penguins <- group_by(penguins, species)

#Notice that the table looks identical, but the metadata notes 'Groups: species [3]' 
print(grouped_penguins)
```
```{r}
#Collapsing the buckets into explicit summary metrics 
species_mass_summary <- summarise(grouped_penguins,
                                  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)
```
The printout summary operation failed and returned an entire column of NA values for certain groups! 
This is an essential behavioural design feature of R known as the Missing Value Trap. If a column contains even a single missing observation flag (NA) any mathematical aggregation function (such as mean(), sd(), or sum()) will automatically return NA to protect you from miscalculating metrics on incomplete data. To override this, we must explicitly declare that missing cells should be dropped during calculation. We'll use the magrittr pipe to wrangle our data, avoiding the missing value trap. 
```{r}
#Overcoming the missing value trap using na.rm = TRUE 
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(), #count total individuals per category 
    mean_mass_g = mean(body_mass_g, na.rm = TRUE), #calculate mean ignoring missing cells 
    sd_mass_g = sd(body_mass_g, na.rm = TRUE) #standard deviation calculation 
  )

print(biological_signal)
```
#1.9 Integrating data grammar with visual diagnostics in qmd 
1.9.1 Wrangling and Plotting in Parallel 
We will combine the pipe (|>) and the basics of ggplot2 into a single reproducible block. 

Challenge: Visualing Body Size and Shape 
Using data loaded, create a code chunk in your .qmd that performs the following 2 tasks simultaneously: 

1. Summary Table: Pipe your data to group_by() and summarise() to calculate the mean body mass for each species and island AND 2. Visualisation: pipe the same dataset into ggplot() to create a boxplot of body_mass_g by species, with the island variable mapped to fill or facet_wrap(). 

```{r}
#penguin_figure_body_mass <- 
  penguins |> 
  group_by(species, island) |> 
  summarise(mean_body_mass_g = mean(body_mass_g, na.rm = TRUE)) |> 
  ggplot(aes(x = species, y = mean_body_mass_g, fill = island)) + 
  geom_col() + 
  theme_minimal()
```
1.9.2 Piping Directly to Visualisation 
Sometimes, you do not need to save a summary table as an object. If your goal is to visualse patterns and uncertainty within your aggregated data, you can extend your pipeline directly into ggplot2. 

Example: Visualising Mean Trends with Uncertainty - instead of creating a summary table object first, you can pipe your grouped data straight into a plot using geom_errorbar() to represent the spread of your data: 

```{r}
#pipe directly from aggregation to plotting with error bars 
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE), 
    sd_mass = sd(body_mass_g, na.rm = TRUE), 
    n = n(), 
    .groups = "drop"
  ) |> 
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) + 
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island", 
       subtitle = "Error bars represent standard deviation", 
       y = "Mean Body Mass (g)", 
       x = "Species") +
  theme_minimal()

mass_compare_plot
```
The above graph is showing that Adelie & Chinstrap penguins have similar mean body masses. It is also showing that Adelie penguins on each of the islands have similar mean body masses. And it is showing that Gentoo penguins have a larger mean body mass compared to the other two penguin species. 

Challenge: Create a new plot. Swap sd_mass (standard deviation) for standard error (sd/sqrt(n)). See how it changes the visual interpretation of the data uncertainty. 

```{r}
#pipe directly from aggregation to plotting with error bars 
mass_sterror_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE), 
    sd_mass = sd(body_mass_g, na.rm = TRUE), 
    n = n(), 
    se_mass = (sd_mass/sqrt(n)),
    .groups = "drop"
  ) |> 
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) + 
  geom_errorbar(aes(ymin = mean_mass - se_mass, 
                    ymax = mean_mass + se_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island", 
       subtitle = "Error bars represent standard error", 
       y = "Mean Body Mass (g)", 
       x = "Species") +
  theme_minimal()

mass_sterror_plot
```
#1.10 Saving, exporting and version milestones 
```{r}
#Create output directories if they do not exist: 
if(!dir.exists("output")) dir.create("output")
if(!dir.exists("output/figures")) dir.create("output/figures") #folder for figures 
if(!dir.exists("output/tables")) dir.create("output/tables") #folder for tables 
if(!dir.exists("Rdata")) dir.create("Rdata") #folder for Rdata objects 
```

TABLES 

```{r}
#1. Exporting our collapsed summary table as a universal flat text file. 
write_csv(biological_signal, "output/tables/penguin_species_mass_summary.csv")

#2. Saving out cleaned morpholigical cohort table as a native R binary file 
saveRDS(penguins_final,"output/tables/penguins_final.rds")
```

FIGURES 

```{r}
ggsave("output/figures/mass_compare_plot.png", 
       plot = mass_compare_plot,
       width = 120, height = 120,
       units = "mm", dpi = 300)
```







